---
title: "Golub Leukemia Dataset Analysis - Impact of Row Scaling (Three-group Classification)"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

-----------------------------------------------------------

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center",
    fig.width = 6,
    fig.height = 6,
    results = "hide")
```

```{r, echo = FALSE}
library(ComplexHeatmap)
library(circlize)
library(GetoptLong)
library(eulerr)
library(cowplot)
```

```{r}
library(golubEsets)
data(Golub_Merge)
m = exprs(Golub_Merge)
colnames(m) = paste0("sample_", colnames(m))
anno = pData(Golub_Merge)
anno = anno[, c("ALL.AML", "T.B.cell")]
anno_col = list("ALL.AML" = c("ALL" = "red", "AML" = "blue"),
	            "T.B.cell" = c("B-cell" = "orange", "T-cell" = "purple"))
```

```{r, echo = FALSE}
library(cola)
res_list_1 = readRDS("/desktop-home/guz/project/development/cola_examples/Golub_leukemia/Golub_leukemia_subgroup.rds")
res_list_2 = readRDS("/desktop-home/guz/project/development/cola_examples/Golub_leukemia/Golub_leukemia_subgroup_unscaled.rds")
# adjust labels 2 -> 1, 1 -> 2 for k = 3
# new_class = res_list_2@consensus_class$`3`$class_df[, "class"]
# new_class[new_class == 1] = -1
# new_class[new_class == 2] = 1
# new_class[new_class == -1] = 2
# res_list_2@consensus_class$`3`$class_df[, "class"] = new_class

# new_membership = res_list_2@consensus_class$`3`$membership
# new_membership = new_membership[, c(2, 1, 3)]
# colnames(new_membership) = paste0("p", 1:3)
# res_list_2@consensus_class$`3`$membership = new_membership

# new_class = res_list_2@consensus_class$`2`$class_df[, "class"]
# new_class[new_class == 1] = -1
# new_class[new_class == 2] = 1
# new_class[new_class == -1] = 2
# res_list_2@consensus_class$`2`$class_df[, "class"] = new_class

# new_membership = res_list_2@consensus_class$`2`$membership
# new_membership = new_membership[, c(2, 1)]
# colnames(new_membership) = paste0("p", 1:2)
# res_list_2@consensus_class$`2`$membership = new_membership

# for(i in seq_along(res_list_2@list)) {
# 	res = res_list_2@list[[i]]
	
# 	new_class = res@object_list$`3`$class_df[, "class"]
# 	new_class[new_class == 1] = -1
# 	new_class[new_class == 2] = 1
# 	new_class[new_class == -1] = 2
# 	res@object_list$`3`$class_df[, "class"] = new_class
	
# 	new_membership = res@object_list$`3`$membership
# 	new_membership = new_membership[, c(2, 1, 3)]
# 	colnames(new_membership) = paste0("p", 1:3)
# 	res@object_list$`3`$membership = new_membership

# 	new_membership_each = res@object_list$`3`$membership_each
# 	new_membership_each[new_membership_each == 1] = -1
# 	new_membership_each[new_membership_each == 2] = 1
# 	new_membership_each[new_membership_each == -1] = 2
# 	res@object_list$`3`$membership_each = new_membership_each

# 	new_class = res@object_list$`2`$class_df[, "class"]
# 	new_class[new_class == 1] = -1
# 	new_class[new_class == 2] = 1
# 	new_class[new_class == -1] = 2
# 	res@object_list$`2`$class_df[, "class"] = new_class
	
# 	new_membership = res@object_list$`2`$membership
# 	new_membership = new_membership[, c(2, 1)]
# 	colnames(new_membership) = paste0("p", 1:2)
# 	res@object_list$`2`$membership = new_membership

# 	new_membership_each = res@object_list$`2`$membership_each
# 	new_membership_each[new_membership_each == 1] = -1
# 	new_membership_each[new_membership_each == 2] = 1
# 	new_membership_each[new_membership_each == -1] = 2
# 	res@object_list$`2`$membership_each = new_membership_each

# 	res_list_2@list[[i]] = res
# }
```

In [the two-group classification](../suppl_10_row_scaling_Golub_2groups/suppl_10_row_scaling_Golub_2groups.html),
AML samples are basically clustered together no matter rows are scaled are
not. Samples can also be separated into
three groups where ALL samples are separated into two subgroups.

### Consensus heatmaps for three subgroups {.tabset}

In the three-group classification, when rows are scaled, only `ATC:skmeans`
generates stable partition (1-PAC score = 0.96), while when rows are not
scaled, no method generates stable partitions, as shown in Table 1 and Figure S11.1.

<style>
table {
  border-collapse: collapse;
}

table, td, th {
  border: 1px solid grey;
}

td, th {
  padding: 2px 8px;
}
</style>

```{r, results = "asis"}
tb1 = get_stats(res_list_1, k = 3)
tb2 = get_stats(res_list_2, k = 3)
tb = data.frame(tb1[, "1-PAC"], tb2[, "1-PAC"])
tb = tb[order(tb[, 1] > 0.9, tb[, 2], decreasing = TRUE), ]
tb[, 1] = round(tb[, 1], 2)
tb[, 2] = round(tb[, 2], 2)

library(kableExtra)
tb[, 1] = cell_spec(tb[, 1], "html", color = ifelse(tb[, 1] >= 0.9, "red", "black"))
tb[, 2] = cell_spec(tb[, 2], "html", color = ifelse(tb[, 2] >= 0.9, "red", "black"))
knitr::kable(tb, col.names = c("Rows are scaled", "Rows are unscaled"),  format = "html", table.attr = "style='width:50%; margin:auto;'", caption = "Table 1. 1-PAC scores of the consensus partitioning under each method (k = 3). Partitions with 1-PAC > 0.9 (numbers in red) are considered as stable partitions.", escape = FALSE)
```

<br>
<br>

#### Rows are scaled

```{r, fig.width = 12, fig.height = 8, fig.cap = "Figure S11.1A. Consensus heatmaps with three groups. Matrix rows are scaled."}
collect_plots(res_list_1, k = 3, anno = anno, anno_col = anno_col)
```

#### Rows are unscaled

```{r, fig.width = 12, fig.height = 8, fig.cap = "Figure S11.1B. Consensus heatmaps with three groups. Matrix rows are unscaled."}
collect_plots(res_list_2, k = 3, anno = anno, anno_col = anno_col)
```

###

We pick the three-group classification from `ATC:skmeans` for the scaled case
because it is the only stable partition, and pick `SD:skmeans` for the
unscaled case because it has the highest 1-PAC score although it is not a
stable partition. Besides the AML/ALL annotation, the original Golub study
also provides B-cell/T-cell annotation for the ALL samples. The cola
classifications and the annotations are visualized in Figure S11.2.

```{r, fig.height = 1, fig.width = 8, fig.cap = "Figure S11.2. Conseusus partitions from ATC:skmeans and SD:skmeans methods."}
cl1 = get_classes(res_list_1["ATC:skmeans"], k = 3)[, 1]
cl2 = get_classes(res_list_2["SD:skmeans"], k = 3)[, 1]

ht = Heatmap(rbind("ATC:skeamns, scaled" = cl1, "SD:skmeans, unscaled" = cl2), name = "Class", col = c("1" = "#66C2A5", "2" = "#FC8D62", "3" = "#8DA0CB"),
	show_row_dend = FALSE, show_column_dend = FALSE,
	top_annotation = HeatmapAnnotation(df = anno, col = anno_col))
draw(ht)
```

To make it easier to describe, we term the classification from the scaled
matrix as $C_{s, 3}$ (or C_{s,3} if it cannot be rendered as math symbols) 
and the classification from the unscaled matrix as $C_{u,3}$ (or C_{u,3}).

In Figure S11.2, we can observe that both classification $C_{s, 3}$ and $C_{u, 3}$
separate ALL samples into two groups, but the classifications are different.
$C_{u,3}$ separates ALL samples into two groups that fit very well with
the B-cell/T-cell annotation while $C_{s, 3}$ gives a new classification.
Note in the Golub study, the B-cell/T-cell classification on ALL samples 
were based on SOM on the unscaled matrix, so it is expected that
$C_{u,3}$ agrees on the B-cell/T-cell classification, also there were
no stability accessment on the classification in the original study.

In the PCA plots in Fgiure 3, all three groups are separated very well
under $C_{s, 3}$, while group 2 and group3 can hardly be separated under
$C_{u, 3}$.

```{r, fig.width = 14, fig.height = 7, fig.cap = "Figure S11.3. PCA plots for visualizing the two-group classification. Classification on the left plot is from $C_{s,3}$ and on the right is from $C_{u,3}$."}
p1 = ~dimension_reduction(res_list_1["ATC:skmeans"], k = 3, method = "PCA")
p2 = ~dimension_reduction(res_list_2["SD:skmeans"], k = 3, method = "PCA", scale_rows = FALSE)
plot_grid(p1, p2, nrow = 1,labels = c("A", "B"))
```


### Signature genes

Next we look at the significantly expressed genes under the two classifications, 
illustrated in Figure S11.4. Both classifications give amount of sigature genes, which
is mainly due to the difference between ALL and AML. To see how different
methods affect the classifiction on ALL samples, we should only look at the ALL samples,
as described in the next section. 

### {.tabset}

#### Rows are scaled

```{r, fig.width = 8, fig.height = 8, out.width = "600px", fig.cap = "Figure S11.4A. Signature genes from C_{s,3}."}
tb1 = get_signatures(res_list_1["ATC:skmeans"], k = 3, anno = anno, anno_col = anno_col, simplify = TRUE)
```

#### Row are unscaled

```{r, fig.width = 8, fig.height = 8, out.width = "600px", fig.cap = "Figure S11.4B. Signature genes from C_{u,3}."}
tb2 = get_signatures(res_list_2["SD:skmeans"], k = 3, anno = anno, anno_col = anno_col, simplify = TRUE, scale_rows = TRUE)
```


### Only look at ALL samples

We remove the samples with class label 1 (which corresponds to AML samples)
in both classifications, and we apply t-test between group 2 and group 3 samples.
The significant genes under the two classifications are visualized as heatmaps
in Figure S11.5.

Since now we only look at the two groups in the two classifications, we modify
the symbols of the two classification to $C^{(2)}_{s,3}$ and $C^{(2)}_{u,3}$.

A direct conclusion from Figure S11.5 is there are far more significant genes under
$C^{(2)}_{s,3}$ (1217 genes) than $C^{(2)}_{u,3}$ (245 genes).

```{r, fig.width = 12, fig.height = 6, fig.cap = "Figure S11.5. Significant genes between group 2 and 3 in the two classifications. Left: from $C^{(2)}_{s,3}$, right: from $C^{(2)}_{u,3}$."}
ind1 = cl1 != 1

ind2 = cl2 != 1

library(genefilter)
mat = get_matrix(res_list_1)
t1 = rowttests(mat[, ind1], factor(cl1[ind1]))
t2 = rowttests(mat[, ind2], factor(cl2[ind2]))

t1$fdr = p.adjust(t1$p.value, "BH")
t2$fdr = p.adjust(t2$p.value, "BH")

col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
anno_col$class = c("1" = "#66C2A5", "2" = "#FC8D62", "3" = "#8DA0CB")
mat1 = mat[t1$fdr < 0.05, ind1]
ht1 = Heatmap(t(scale(t(mat1))), name = "Expression", 
	show_row_names = FALSE, col = col_fun,
	show_column_names = FALSE, show_row_dend = FALSE, show_column_dend = FALSE,
	top_annotation = HeatmapAnnotation(class = cl1[ind1], T.B.cell = anno$T.B.cell[ind1], col = anno_col),
	column_split = cl1[ind1], row_split = ifelse(t1$statistic[t1$fdr < 0.05] > 0, "group1", "group2"),
	column_title = qq("scaled, @{sum(t1$fdr < 0.05)} genes with FDR < 0.05"))
p1 = grid.grabExpr(ht1 <- draw(ht1, merge_legends = TRUE))
od1 = row_order(ht1)

mat2 = mat[t2$fdr < 0.05, ind2]
ht2 = Heatmap(t(scale(t(mat2))), name = "Expression", 
	show_row_names = FALSE, col = col_fun,
	show_column_names = FALSE, show_row_dend = FALSE, show_column_dend = FALSE,
	top_annotation = HeatmapAnnotation(class = cl2[ind2], T.B.cell = anno$T.B.cell[ind2], col = anno_col),
	column_split = cl2[ind2], row_split = ifelse(t2$statistic[t2$fdr < 0.05] > 0, "group1", "group2"),
	column_title = qq("unscaled, @{sum(t2$fdr < 0.05)} genes with FDR < 0.05"))
p2 = grid.grabExpr(ht2 <- draw(ht2, merge_legends = TRUE))
od2 = row_order(ht2)

plot_grid(p1, p2, nrow = 1, labels = c("A", "B"))
```

Since Figure S11.5 only shows the relative expression (the scaled expression), in Figure S11.6,
we check the distribution of the absolute expression level of the signature genes
in the two classifications, visualized in Figure S11.6. Figure S11.6 shows both signature genes
are more enriched in the highly expressed genes, where signatures from $C^{(2)}_{s,3}$
also cover genes with moderate expression levels.

```{r, fig.width = 8, fig.height = 6, fig.cap = "Figure S11.6. Distribution of the absolute expression of signature genes. A and C: genes are ordered based on their base mean (mean expression among all samples) where significant genes are marked in blue and red. B and D: number of signatures that fall in every 20-gene interval, the same ordering on x-axes as Figure A and C."}
par(mfrow = c(2, 2), mar = c(2, 4, 4, 1))
s = rowMeans(get_matrix(res_list_1))
s_rk = structure(rank(s), names = names(s))
plot(s_rk, s, type = "h", col = "grey", xlim = c(0, 4200), xlab = "", ylab = "Base mean", main = "A) scaled, genes are ordered by base mean")
nm = rownames(mat2)[ od2[[1]] ]
lines(s_rk[nm], s[nm], col = "blue", type = "h")
hist(s_rk[nm], nclass = 21, xlim = c(0, 4200), xlab = "", main = "B) scaled, #signatures in every 20-gene bins")

plot(s_rk, s, type = "h", col = "grey", xlim = c(0, 4200), xlab = "", ylab = "Base mean", main = "C) unscaled, genes are ordered by base mean")
nm = rownames(mat2)[ od2[[2]] ]
lines(s_rk[nm], s[nm], col = "red", type = "h")
hist(s_rk[nm], nclass = 21, xlim = c(0, 4200), xlab = "", main = "D) unscaled, #signatures in every 20-gene bins")
```

The two sets of signature genes have very small overlap, illustrated in Figure S11.7.

```{r, fig.width = 8, fig.height = 3, fig.cap = "Figure S11.7. Overlap of signature genes in $C^{(2)}_{s,3}$ and $C^{(2)}_{u,3}$."}
plot(euler(list(scaled = which(t1$fdr < 0.05), unscaled = which(t2$fdr < 0.05))), quantities = TRUE)
```

Similar as the analysis in [the two-group classification](../suppl_10_row_scaling_Golub_2groups/suppl_10_row_scaling_Golub_2groups.html), we check the biological functions that are enriched
in the two sets of signature genes from $C^{(2)}_{s,3}$ and $C^{(2)}_{u,3}$. Mainly because of
the high amount of signature genes from $C^{(2)}_{s,3}$, it has far more enriched GO functions
than $C^{(2)}_{u,3}$ (Figure S11.8).

```{r}
library(hu6800.db)
x = hu6800ENTREZID
mapped_probes = mappedkeys(x)
id_mapping = unlist(as.list(x[mapped_probes]))
tb1 = functional_enrichment(rownames(mat1)[unlist(od1)], id_mapping = id_mapping, 
	ontology = "BP", verbose = FALSE)[[1]]
tb1 = tb1[tb1$Count >= 5, , drop = FALSE]
tb2 = functional_enrichment(rownames(mat1)[unlist(od2)], id_mapping = id_mapping, 
	ontology = "BP", verbose = FALSE)[[1]]
tb2 = tb2[tb2$Count >= 5, , drop = FALSE]
```    

```{r, fig.width = 12, fig.height = 7, fig.cap = "Figure S11.8. Gene ontology enrichment on the two sets of signature genes illustrated in Figure S11.5."}
ago = c(rownames(tb1), rownames(tb2))
ago = unique(ago)
pm = matrix(1, nrow = length(ago), ncol = 2)
rownames(pm) = ago
colnames(pm) = c("C^{(2)}_{s,3}", "C^{(2)}_{u,3}")
pm[tb1$ID, 1] = tb1$p.adjust
pm[tb2$ID, 2] = tb2$p.adjust

fdr_cutoff = 0.01
pm = pm[apply(pm, 1, function(x) any(x < fdr_cutoff)), ]
all_go_id = rownames(pm)
col_fun_p = colorRamp2(c(0, -log10(fdr_cutoff), 4), c("green", "white", "red"))
ht_fdr = Heatmap(-log10(pm), col = col_fun_p, name = "FDR",
	show_row_names = FALSE, cluster_columns = FALSE,
	border = "black", column_title = "FDR",
	heatmap_legend_param = list(at = c(0, -log10(fdr_cutoff), 4), 
		labels = c("1", fdr_cutoff, "<0.0001")),
	width = unit(1.5, "cm"))

library(simplifyEnrichment)
sim_mat = GO_similarity(all_go_id)
pm = pm[rownames(sim_mat), , drop = FALSE]
col_fun_p = colorRamp2(c(0, -log10(fdr_cutoff), 4), c("green", "white", "red"))
ht_fdr = Heatmap(-log10(pm), col = col_fun_p, name = "FDR",
	show_row_names = FALSE, cluster_columns = FALSE,
	border = "black", column_title = "FDR",
	heatmap_legend_param = list(at = c(0, -log10(fdr_cutoff), 4), 
		labels = c("1", fdr_cutoff, "<0.0001")),
	width = unit(1.5, "cm"),
	column_labels = gt_render(c("*C*<sup>(2)</sup><sub>s,3</sub>", "*C*<sup>(2)</sup><sub>u,3</sub>")))

invisible(simplifyGO(sim_mat, ht_list = ht_fdr, word_cloud_grob_param = list(max_width = 120), verbose = FALSE))
```


### Conclusion

As conclusions, it seems the classification from $C^{(2)}_{s,3}$ is more reasonable because
there are many signature genes as well as enriched biolofical functions, which might
provide a new and different view to the B-cell/T-cell classification.


### Gene ontology enrichment {.tabset}

#### $C^{(2)}_{s,3}$  (`r sum(tb1$p.adjust <= fdr_cutoff)` terms)

```{r, results = "markup"}
# tb1 = functional_enrichment(rownames(mat1)[od1[[1]]], id_mapping = id_mapping, 
# 	ontology = "Reactome", organism = "human", universe = universe, verbose = FALSE)[[1]]
tb1$qvalue = NULL
tb1$geneID = NULL
# tb1$Description = qq("<a href='https://reactome.org/content/detail/@{tb1$ID}' target='_blank'>@{tb1$Description}</a>", collapse = FALSE)
# tb1$ID = NULL
# tb1 = tb1[tb1$p.adjust <= fdr_cutoff & tb1$Count >= 5, , drop = FALSE]
# gene_list = strsplit(tb1$geneID, "/")
# tb1$geneID = NULL

# n = length(gene_list)
# jmat = matrix(1, nrow = n, ncol = n)
# for(i in 1:(n-1)) {
# 	for(j in (i+1):n) {
# 		jmat[i, j] = jmat[j, i] = length(intersect(gene_list[[i]], gene_list[[j]]))/length(union(gene_list[[i]], gene_list[[j]]))
# 	}
# }
# Heatmap(jmat)
kable(tb1[tb1$p.adjust <= fdr_cutoff, , drop = FALSE], digits = 4, row.names = FALSE)
```

#### $C^{(2)}_{u,3}$ (`r sum(tb2$p.adjust <= fdr_cutoff)` terms)

```{r, results = "markup"}
# tb2 = functional_enrichment(rownames(mat1)[od1[[2]]], id_mapping = id_mapping, 
	# ontology = "Reactome", organism = "human", universe = universe, verbose = FALSE)[[1]]
tb2$qvalue = NULL
# tb2$Description = qq("<a href='https://reactome.org/content/detail/@{tb2$ID}' target='_blank'>@{tb2$Description}</a>", collapse = FALSE)
# tb2$ID = NULL
tb2$geneID = NULL
kable(tb2[tb2$p.adjust <= fdr_cutoff, , drop = FALSE], digits = 4, row.names = FALSE)
```

###

<br>
<br>
<br>
<br>

<style>
#gene-ontology-enrichment div {
	max-height: 400px;
	overflow-y: auto;
}
#gene-ontology-enrichment table td {
	padding: 2px 4px;
}
#gene-ontology-enrichment table th {
	padding: 2px 4px;
}
</style>



