---
title: "Jaccard Index"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

-----------------------------------------

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    fig.align = "center")
```


Following two consensus heatmaps (Figure 1) are for $k$ = 2 and 3 (2 and 3
groups) on
[GDS1875](https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS1875) GDS
dataset under top-value method **ATC** and partitioning method **hclust**
(cola report is [here](https://cola-gds.github.io/GDS1875/cola_report.html)).
Figure 1 illustrates that when the number of subgroups increases form 2 to 3,
a tiny subgroup only with one single sample is separated from one of the big
subgroups.

```{r, fig.width = 12, fig.height = 6, echo = FALSE, fig.align = "center", fig.cap = "Figure 1. Consensus heatmaps for k = 2 and 3."}
library(circlize)
library(ComplexHeatmap)
library(GetoptLong)
library(cowplot)

id = "GDS1875"
res = readRDS(qq("/icgc/dkfzlsdf/analysis/B080/guz/cola_test/GDS/@{id}/@{id}_cola_all.rds"))
res = res["ATC:hclust"]

p1 = grid.grabExpr(consensus_heatmap(res, k = 2, anno = NULL))
p2 = grid.grabExpr(consensus_heatmap(res, k = 3, anno = NULL))

p = plot_grid(p1, p2, nrow = 1)
print(p)
```

Next we plot the 1-PAC scores as well Jaccard indices for the partitions under each $k$ (Figure 2).
We can see the 1-PAC scores are both very high at $k$ = 2 and 3, which means the partitions are both
very stable for the two $k$. However, the Jaccard index is extremely high at $k$ =  3 (0.97), 
which quantitatively shows partitions in $k$ = 3 is highly similar as $k$ = 2, and on the other hand,
an extra subgroup only with one single sample does not provide any useful output for the analysis. Thus,
we decline $k$ = 3 is a better choice while still take $k$ = 2 as the best number of subgroups.

```{r, fig.width = 10, fig.height = 5, echo = FALSE, fig.align = "center", fig.cap = "Figure 2. The 1-PAC scores as well as Jaccard indices for the partitions under each k."}
stat_df = as.data.frame(get_stats(res))
k = 3

par(mfrow = c(1, 2))
plot(stat_df$k, stat_df$`1-PAC`, main = "1-PAC", xlab = "k", ylab = "1-PAC", 
	type = "o")
plot(stat_df$k, stat_df$Jaccard, main = "Jaccard index", xlab = "k", ylab = "Jaccard index",
	type = "o")
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

