---
title: "Compare Top-value Methods"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

-----------------------------------------------------------

```{r, echo = FALSE, message = FALSE}
library(cola)
```

We compare the top $n$ ($n$ = 500, 1000, 2000) rows scored by four methods: SD
(standard deviation), CV (coefficient of variation), MAD (median absolute
deviation) and ATC (ability to correlate to other rows) with five datasets:
[Golub leukemia
dataset](https://jokergoo.github.io/cola_examples/Golub_leukemia/), [Ritz ALL
dataset](https://jokergoo.github.io/cola_examples/Ritz_ALL) and [TCGA GBM
microarray dataset](https://jokergoo.github.io/cola_examples/TCGA_GBM), [HSMM
single cell RNASeq
dataset](https://jokergoo.github.io/cola_examples/HSMM_single_cell) and [MCF10CA
single cell RNASeq
dataset](https://jokergoo.github.io/cola_examples/MCF10CA_scRNAseq). For
each dataset, we visualize the top rows (or, in other words, the top genes) by
heatmaps (where rows are scaled by $z$-score method) and Euler diagram.

### Golub leukemia dataset {.tabset}

```{r, echo = FALSE}
res_list = readRDS("/desktop-home/guz/project/development/cola_examples/Golub_leukemia/Golub_leukemia_subgroup.rds")
```

#### top n = 500

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.1A. Heatmap of the top 500 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_heatmap(res_list, top_n = 500, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.2A. Euler diagram of the top 500 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_overlap(res_list, top_n = 500, method = "euler")
```

#### top n = 1000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.1B. Heatmap of the top 1000 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_heatmap(res_list, top_n = 1000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.2B. Euler diagram of the top 1000 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_overlap(res_list, top_n = 1000, method = "euler")
```

#### top n = 2000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.1C. Heatmap of the top 2000 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_heatmap(res_list, top_n = 2000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.2C. Euler diagram of the top 2000 genes with the highest SD, CV, MAD and ATC scores, Golub leukemia dataset."}
top_rows_overlap(res_list, top_n = 2000, method = "euler")
```


### Ritz ALL dataset {.tabset}

```{r, echo = FALSE}
res_list = readRDS("/desktop-home/guz/project/development/cola_examples/Ritz_ALL/Ritz_ALL_subgroup.rds")
```


#### top n = 500

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.7A. Heatmap of the top 500 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_heatmap(res_list, top_n = 500, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.8A. Euler diagram of the top 500 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_overlap(res_list, top_n = 500, method = "euler")
```

#### top n = 1000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.7B. Heatmap of the top 1000 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_heatmap(res_list, top_n = 1000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.8B. Euler diagram of the top 1000 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_overlap(res_list, top_n = 1000, method = "euler")
```

#### top n = 2000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.7C. Heatmap of the top 2000 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_heatmap(res_list, top_n = 2000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.8C. Euler diagram of the top 2000 genes with the highest SD, CV, MAD and ATC scores, Ritz ALL dataset."}
top_rows_overlap(res_list, top_n = 2000, method = "euler")
```


### TCGA GBM microarray dataset {.tabset}


```{r, echo = FALSE}
res_list = readRDS("/desktop-home/guz/project/development/cola_examples/TCGA_GBM/TCGA_GBM_subgroup.rds")
```


#### top n = 500

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.9A. Heatmap of the top 500 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_heatmap(res_list, top_n = 500, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.10A. Euler diagram of the top 500 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_overlap(res_list, top_n = 500, method = "euler")
```

#### top n = 1000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.9B. Heatmap of the top 1000 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_heatmap(res_list, top_n = 1000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.10B. Euler diagram of the top 1000 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_overlap(res_list, top_n = 1000, method = "euler")
```

#### top n = 2000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.9C. Heatmap of the top 2000 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_heatmap(res_list, top_n = 2000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.10C. Euler diagram of the top 2000 genes with the highest SD, CV, MAD and ATC scores, TCGA GBM microarray dataset."}
top_rows_overlap(res_list, top_n = 2000, method = "euler")
```


### HSMM single cell RNASeq dataset {.tabset}

```{r, echo = FALSE}
res_list = readRDS("/desktop-home/guz/project/development/cola_examples/HSMM_single_cell/HSMM_single_cell_subgroup.rds")
```

#### top n = 500

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.3A. Heatmap of the top 500 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 500, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.4A. Euler diagram of the top 500 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 500, method = "euler")
```

#### top n = 1000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.3B. Heatmap of the top 1000 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 1000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.4B. Euler diagram of the top 1000 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 1000, method = "euler")
```

#### top n = 2000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.3C. Heatmap of the top 2000 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 2000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.4C. Euler diagram of the top 2000 genes with the highest SD, CV, MAD and ATC scores, HSMM single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 2000, method = "euler")
```

### MCF10CA single cell RNASeq dataset {.tabset}

```{r, echo = FALSE}
res_list = readRDS("/desktop-home/guz/project/development/cola_examples/MCF10CA_scRNAseq/MCF10CA_scRNAseq_subgroup.rds")
```


#### top n = 500

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.5A. Heatmap of the top 500 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 500, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.6A. Euler diagram of the top 500 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 500, method = "euler")
```

#### top n = 1000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.5B. Heatmap of the top 1000 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 1000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.6B. Euler diagram of the top 1000 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 1000, method = "euler")
```

#### top n = 2000

```{r, echo = FALSE, fig.width = 10, fig.height = 2.5, fig.align = "center", fig.cap = "Figure S2.5C. Heatmap of the top 2000 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_heatmap(res_list, top_n = 2000, anno = NULL)
```

```{r, echo = FALSE, fig.width = 3, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.6C. Euler diagram of the top 2000 genes with the highest SD, CV, MAD and ATC scores, MCF10CA single cell RNASeq dataset."}
top_rows_overlap(res_list, top_n = 2000, method = "euler")
```



### Uniqueness of the top rows

```{r, echo = FALSE}
library(GetoptLong)
calc_uniqueness = function(project) {
	if(project == "recount2") {
		id_list = dir("/icgc/dkfzlsdf/analysis/B080/guz/cola_test/recount2")
	} else if(project == "GDS") {
		id_list = dir("/icgc/dkfzlsdf/analysis/B080/guz/cola_test/GDS", pattern = "^GDS\\d+$")
	}

	df = file.info(qq("/icgc/dkfzlsdf/analysis/B080/guz/cola_test/@{project}/@{id_list}", collapse = FALSE))
	id_list = basename(rownames(df)[df$isdir])

	frac_df = NULL
	basemean_quan_lt = list()
	sim_list = list()
	for(i in seq_along(id_list)) {
		id = id_list[i]
		qqcat("loading @{id}...\n")
		res = readRDS(qq("/icgc/dkfzlsdf/analysis/B080/guz/cola_test/@{project}/@{id}/@{id}_cola_all.rds"))
		all_top_value_list = res@.env$all_top_value_list[c("SD", "CV", "MAD", "ATC")]
		lt = lapply(all_top_value_list, function(x) order(x, decreasing = TRUE)[1:1000])

		frac_df = rbind(frac_df, matrix(
			c("SD" = length(setdiff(lt[[1]], unlist(lt[-1]))),
			  "CV" = length(setdiff(lt[[2]], unlist(lt[-2]))),
			  "MAD" = length(setdiff(lt[[3]], unlist(lt[-3]))),
			  "ATC" = length(setdiff(lt[[4]], unlist(lt[-4])))
			  ), nrow = 1))

		m = get_matrix(res)
		basemean_quan = rank(rowMeans(m))/nrow(m)
		basemean_quan_lt[[i]] = list(
			SD = basemean_quan[ lt$SD ],
			CV = basemean_quan[ lt$CV ],
			MAD = basemean_quan[ lt$MAD ],
			ATC = basemean_quan[ lt$ATC ]
		)

		sim_list[[i]] = matrix(1, nrow = 4, ncol = 4)
		rownames(sim_list[[i]]) = colnames(sim_list[[i]]) = c("SD", "CV", "MAD", "ATC")
		sim_list[[i]]["SD", "CV"] = sim_list[[i]]["CV", "SD"] = length(intersect(lt[[1]], lt[[2]]))/1000
		sim_list[[i]]["SD", "MAD"] = sim_list[[i]]["MAD", "SD"] = length(intersect(lt[[1]], lt[[3]]))/1000
		sim_list[[i]]["SD", "ATC"] = sim_list[[i]]["ATC", "SD"] = length(intersect(lt[[1]], lt[[4]]))/1000
		sim_list[[i]]["CV", "MAD"] = sim_list[[i]]["MAD", "CV"] = length(intersect(lt[[2]], lt[[3]]))/1000
		sim_list[[i]]["CV", "ATC"] = sim_list[[i]]["ATC", "CV"] = length(intersect(lt[[2]], lt[[4]]))/1000
		sim_list[[i]]["MAD", "ATC"] = sim_list[[i]]["ATC", "MAD"] = length(intersect(lt[[3]], lt[[4]]))/1000

	}
	colnames(frac_df) = c("SD", "CV", "MAD", "ATC")

	return(list(frac_df = frac_df, basemean_quan_lt = basemean_quan_lt, sim_list = sim_list))
}

if(file.exists("top_rows_uniqueness_GDS.rds")) {
	uniqueness_GDS = readRDS("top_rows_uniqueness_GDS.rds")
} else {
	uniqueness_GDS = calc_uniqueness("GDS")
	saveRDS(uniqueness_GDS, file = "top_rows_uniqueness_GDS.rds")
}

if(file.exists("top_rows_uniqueness_recount2.rds")) {
	uniqueness_recount2 = readRDS("top_rows_uniqueness_recount2.rds")
} else {
	uniqueness_recount2 = calc_uniqueness("recount2")
	saveRDS(uniqueness_recount2, file = "top_rows_uniqueness_recount2.rds")
}
```

We calculate the uniqueness of the top 1000 rows (genes) by the four top-value
methods for GDS cohort (`r nrow(uniqueness_GDS$frac)` datasets) and recount2
cohort (`r nrow(uniqueness_recount2$frac)` datasets). The uniqueness for a
top-value method is calculated as the fraction of the top 1000 genes that are
not in the top 1000 genes by any of the other three methods. E.g., for one
dataset, the uniqueness of ATC method is calculated as:

```{r, eval = FALSE}
length(setdiff(S_ATC, union(S_SD, union(S_CV, S_MAD))))/1000
```

where `S_ATC`, `S_SD`, `S_CV` and `S_MAD` are the sets of top 1000 genes under each method.


```{r, echo = FALSE}
options(digits = 3)
```

In the following boxplots (Figure S2.11), we find for the GDS datasets which are microarray
datasets, generally, ATC method extracts top genes which are more unique
compared to other three methods (mean uniqueness, SD: `r mean(uniqueness_GDS$frac_df[, "SD"])/1000`, 
CV: `r mean(uniqueness_GDS$frac_df[, "CV"])/1000`, MAD: `r mean(uniqueness_GDS$frac_df[,"MAD"])/1000`, 
ATC: `r mean(uniqueness_GDS$frac_df[, "ATC"])/1000`). For recount2 datasets which
are RNASeq datasets, CV method also extracts quite large fraction of unique
top genes (mean uniqueness, SD: `r mean(uniqueness_recount2$frac_df[, "SD"])/1000`, 
CV: `r mean(uniqueness_recount2$frac_df[, "CV"])/1000`, MAD: `r mean(uniqueness_recount2$frac_df[, "MAD"])/1000`, 
ATC: `r mean(uniqueness_recount2$frac_df[, "ATC"])/1000`). Since RNASeq can
measure genes which have very low expression (according to [the recount2 pipeline](https://jhubiostatistics.shinyapps.io/recount/)), we guess the high fraction of
CV-unique top genes is due to the lowly expressed genes in the recount2
datasets (recall CV is defined as the standard deviation dividing the mean
where small mean can give large CV value).

```{r, echo = FALSE, fig.align = "center", fig.cap = "Figure S2.11. Uniqueness of top-value methods in GDS datasets and recount2 datasets."}
par(mfrow = c(1, 2), mar = c(2.1, 4.1, 5.1, 1.1))
boxplot(as.data.frame(uniqueness_GDS$frac_df/1000), ylim = c(0, 1), 
	ylab = "Uniqueness", main = "GDS datasets")
boxplot(as.data.frame(uniqueness_recount2$frac_df/1000), ylim = c(0, 1), 
	ylab = "Uniqueness", main = "recount2 datasets")
```


Next we check the base mean for the top genes (The base mean is the mean
absolute expression level for genes). To make the base mean comparable among
datasets, the base mean values are replaced by the corresponding rank
normalized by the total number of genes in that dataset, (calculated as
`rank(base_mean)/length(base_mean)`). For each top-value method in each
dataset, the mean rank for the top 1000 genes is used to measure the average
base expression level.

In the following boxplots in Figure S2.12, we see very clearly that in recount2
datasets, the top 1000 genes by CV method have much lower expression than the
top 1000 genes by other top-value methods.

```{r, echo = FALSE, fig.align = "center", fig.cap = "Figure S2.12. Base expression level of top 1000 genes in GDS and recount2 datasets."}
par(mfrow = c(1, 2), mar = c(2.1, 4.1, 5.1, 1.1))
lt = list(
	SD = sapply(uniqueness_GDS$basemean_quan_lt, function(x) mean(x$SD)),
	CV = sapply(uniqueness_GDS$basemean_quan_lt, function(x) mean(x$CV)),
	MAD = sapply(uniqueness_GDS$basemean_quan_lt, function(x) mean(x$MAD)),
	ATC = sapply(uniqueness_GDS$basemean_quan_lt, function(x) mean(x$ATC))
)
boxplot(lt, ylim = c(0, 1), ylab = "Relative rank of base expression", main = "GDS datasets")
lt = list(
	SD = sapply(uniqueness_recount2$basemean_quan_lt, function(x) mean(x$SD)),
	CV = sapply(uniqueness_recount2$basemean_quan_lt, function(x) mean(x$CV)),
	MAD = sapply(uniqueness_recount2$basemean_quan_lt, function(x) mean(x$MAD)),
	ATC = sapply(uniqueness_recount2$basemean_quan_lt, function(x) mean(x$ATC))
)
boxplot(lt, ylim = c(0, 1), ylab = "Relative rank of base expression", main = "recount2 datasets")
```

Figure S2.13 visualizes the average pair-wise overlap among datasets where we can
also see the uniqueness of ATC. For $n$ datasets in a cohort, the value of the
mean overlap between method $i$ and $j$ is defined as:

$$\frac{1}{n}\sum_k^n{p_{i,j,k}}$$

where $p_{ijk}$ is the overlap between method $i$ and $j$ in dataset $k$ and
is defined as:

$$p_{i,j,k} = \frac{| S_{i,k} \bigcap S_{j,k} |}{1000}$$

where $S_{ik}$ and $S_{jk}$ are the sets of top 1000 genes extracted by method
$i$ and $j$ in dataset $k$.

```{r, echo = FALSE, message = FALSE, fig.width = 6, fig.height = 3, fig.align = "center", fig.cap = "Figure S2.13. The average pair-wise overlap among datasets."}
m1 = matrix(0, nrow = 4, ncol = 4)
for(i in seq_along(uniqueness_GDS$sim_list)) {
	m1 = m1 + uniqueness_GDS$sim_list[[i]]
}
m1 = m1/length(uniqueness_GDS$sim_list)

m2 = matrix(0, nrow = 4, ncol = 4)
for(i in seq_along(uniqueness_recount2$sim_list)) {
	m2 = m2 + uniqueness_recount2$sim_list[[i]]
}
m2 = m2/length(uniqueness_recount2$sim_list)

library(cowplot)
library(ComplexHeatmap)
ht1 = Heatmap(m1, name = "overlap", column_title = "GDS datasets", show_row_dend = FALSE, show_column_dend = FALSE,
	cell_fun = function(j, i, x, y, w, h, fill) {
		grid.text(round(m1[i, j], 2), x, y, gp = gpar(fontsize = 8))
	})
p1 = grid::grid.grabExpr(draw(ht1))
ht2 = Heatmap(m2, name = "overlap", column_title = "recount2 datasets", show_row_dend = FALSE, show_column_dend = FALSE,
	cell_fun = function(j, i, x, y, w, h, fill) {
		grid.text(round(m2[i, j], 2), x, y, gp = gpar(fontsize = 8))
	})
p2 = grid::grid.grabExpr(draw(ht2))
plot_grid(p1, p2, nrow = 1)
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

