---
title: "The ATC Method"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

-------------------------------------------

```{r, echo = FALSE, message = FALSE}
library(matrixStats)
library(mvtnorm)
library(ComplexHeatmap)
test_ATC = function(nr1, nr2, nr3, cor1, cor2, cor3, min_cor = 0, beta = 1) {
	
	sigma = matrix(cor1, nrow = nr1, ncol = nr1); diag(sigma) = 1
	mat1 = t(rmvnorm(100, mean = rep(0, nr1), sigma = sigma))

	sigma = matrix(cor2, nrow = nr2, ncol = nr2); diag(sigma) = 1
	mat2 = t(rmvnorm(100, mean = rep(0, nr2), sigma = sigma))

	sigma = matrix(cor3, nrow = nr3, ncol = nr3); diag(sigma) = 1
	mat3 = t(rmvnorm(100, mean = rep(0, nr3), sigma = sigma))

	mat = t(rbind(mat1, mat2, mat3))

	grid.newpage()
	par(mfrow = c(1, 4))
	plot.new()
	par(mar = c(5.1, 5, 2, 1), cex = 0.8)
	v = NULL
	plot(NULL, xlim = c(0, 1), ylim = c(0, 1), ylab = "P(X <= x)", xlab = "absolute correlation")
	for(i in 1:ncol(mat)) {
	    cor_v <- abs(cor(mat[, i, drop = FALSE], mat[, -i, drop = FALSE]))^beta
	    cor_v = sort(cor_v)
	    f = ecdf(cor_v)
	    
	    cor_v = seq(min_cor, 1, length = 100)
	    n2 = length(cor_v)
	    v[i] = sum((cor_v[2:n2] - cor_v[1:(n2-1)])*f(cor_v[-n2]))
	    x = seq(0, 1, length = 100)
	    lines(x, f(x), col = ifelse(i <= nr1, "#00000080", ifelse(i <= nr1+nr2, "#FF000080", "#00FF0080")))
	    if(min_cor > 0) abline(v = min_cor, lty = 3, col = "grey")
	}
	legend("bottomright", lty = 1, col = 1:3, legend = c("group1", "group2", "group3"))
	v = (1 - min_cor) - v

	dx_list = NULL
	dy_list = NULL
	for(i in 1:ncol(mat)) {
	    cor_v <- abs(cor(mat[, i, drop = FALSE], mat[, -i, drop = FALSE]))
	    d = density(cor_v)
	    dx_list = c(dx_list, list(d$x))
	    dy_list = c(dy_list, list(d$y))
	}
	max_y = max(unlist(dy_list))


	plot(v, pch = 16, xlab = "row index", ylab = "ATC", 
		col = ifelse(seq_along(v) <= nr1, "#000000", ifelse(seq_along(v) <= nr1+nr2, "#FF0000", "#00FF00")))

	plot(colSds(mat), pch = 16, xlab = "row index", ylab = "SD", col = ifelse(seq_along(v) <= nr1, "#000000", ifelse(seq_along(v) <= nr1+nr2, "#FF0000", "#00FF00")))

	pushViewport(viewport(x = 0, y = 1, width = 0.25, height = 1, just = c("left", "top")))
	library(ComplexHeatmap)
	split = factor(c(rep(1, nr1), rep(2, nr2), rep(3, nr3)), levels = c(1, 2, 3))
	ht = Heatmap(t(mat), row_split = split, cluster_row_slices = FALSE, row_title = NULL,
		show_row_dend = FALSE, show_column_dend = FALSE, show_heatmap_legend = FALSE,
		left_annotation = rowAnnotation(split = split, col = list(split = c("1" = 1, "2" = 2, "3" = 3)),
			annotation_legend_param = list(
				split = list(labels = c("group1", "group2", "group3"), nrow = 1, title = "")),
			show_annotation_name = FALSE))
	draw(ht, newpage = FALSE, padding = unit(c(2, 10, 5, 5), "mm"), heatmap_legend_side = "bottom")
	popViewport()
}
```

For row $i$ in a matrix $\mathbf{M}$, $\mathbf{x}$ is a vector of**the absolute
correlation** to all other rows, the ATC (ability to correlate to other rows)
score for row $i$ is defined as:

$$ATC_i = 1 - \int_0^1F(x)dx$$

where $F(x)$ is the cumulative distribution function (CDF) of $x$.

Figure 1 shows the empirical CDF (ECDF) curve of $\mathbf{x}$ for 
a certain row $i$. The ATC score corresponds to the red area above the ECDF curve. It can be
imagined that when row $i$ correlates stronger with other rows, the ECDF curve
shifts more to the right, thus with higher ATC scores.

```{r, echo = FALSE, fig.width = 5, fig.height = 5, fig.align = "center", fig.cap = "Figure 1. Demonstration of the ATC score"}
par(mar = c(4, 4, 1, 1))
cor_v = c(0.00296, 0.00323, 0.00493, 0.00511, 0.00513, 0.00653, 0.00838, 0.00859, 0.00923, 0.0116, 0.0141, 0.0142, 0.0144, 0.0161, 0.0186, 0.0195, 0.0205, 0.0222, 0.0231, 0.0237, 0.0242, 0.025, 0.0255, 0.0277, 0.028, 0.029, 0.029, 0.0314, 0.0322, 0.0355, 0.0357, 0.0363, 0.0379, 0.0392, 0.0411, 0.0423, 0.0426, 0.0441, 0.0449, 0.046, 0.0501, 0.0508, 0.0529, 0.0536, 0.054, 0.0541, 0.0549, 0.0567, 0.0576, 0.0581, 0.0605, 0.0608, 0.0639, 0.0671, 0.0682, 0.0698, 0.0717, 0.0737, 0.0747, 0.075, 0.0773, 0.0816, 0.0834, 0.0837, 0.0869, 0.0899, 0.0928, 0.0937, 0.0945, 0.0946, 0.0964, 0.0981, 0.099, 0.101, 0.102, 0.102, 0.103, 0.104, 0.106, 0.109, 0.111, 0.115, 0.115, 0.116, 0.116, 0.121, 0.124, 0.127, 0.127, 0.127, 0.133, 0.135, 0.135, 0.135, 0.138, 0.143, 0.144, 0.146, 0.156, 0.16, 0.164, 0.169, 0.173, 0.178, 0.179, 0.179, 0.194, 0.202, 0.211, 0.253, 0.396, 0.429, 0.447, 0.455, 0.457, 0.465, 0.466, 0.477, 0.487, 0.488, 0.498, 0.518, 0.519, 0.522, 0.524, 0.526, 0.528, 0.539, 0.54, 0.541, 0.544, 0.546, 0.548, 0.551, 0.552, 0.556, 0.558, 0.559, 0.565, 0.568, 0.568, 0.574, 0.574, 0.576, 0.576, 0.579, 0.58, 0.582, 0.585, 0.59, 0.591, 0.592, 0.594, 0.596, 0.597, 0.609, 0.614, 0.615, 0.628)

cor_v = sort(cor_v)
f = ecdf(cor_v)
x = seq(0, 1, length = 100)
y = f(x)

plot(NULL, xlim = c(0, 1), ylim = c(0, 1), xlab = "x = |correlation|", ylab = "P(X <= x)")
x2 = seq(0, 1, length = 100)
polygon(c(x2, rev(x2)), c(f(x2), rep(1, length(x2))), col = "#FF000040", border = NA)
polygon(c(x2, rev(x2)), c(f(x2), rep(0, length(x2))), col = "#00FF0040", border = NA)
lines(x, y)
text(0.2, 0.9, "ATC score")
```


A simulation test is performed to demonstrate the attributes of the ATC method. A
matrix with 160 rows, 100 columns with random values drawn from a multivariate
normal distribution is generated. The 160 rows are configured as follows:

1. 100 rows with mean of 0. The covariance matrix is set to 0 and the
   diagonal with 1;
2. 10 rows with mean of 0. The covariance matrix is set to 0.8 and the
   diagonal with 1. This simulates high correlations but in a small group of rows;
3. 50 rows with mean of 0. The covariance matrix is set to 0.5 and the
   diagonal with 1. This simulates intermediate correlations but in a large group of
   rows.

```{r, echo = FALSE, fig.width = 14, fig.height = 14/4, fig.align = "center", fig.cap = "Figure 2. A simulation test with three groups of rows with different correlation structures"}
set.seed(12345)
test_ATC(100, 10, 50, 0, 0.8, 0.5)
```

The first plot in Figure 2 is the heatmap for the random matrix, split
by the three groups of rows. In the second plot, they are ECDF curves of the
correlation of the 160 rows. The third plot is the ATC scores
for all 160 rows and the fourth plot is the standard deviation for the 160
rows. Different colors represent different row groups.

All the 160 rows have a similar variance of 1 that they cannot be distinguished
by using variance. As a contrast, the rows with non-zero covariance have
higher ATC scores (the red and green groups). The ATC scores are even higher
when the number of correlated rows increases, although the correlation value
itself is relatively small (the green group). This shows ATC method can assign
higher scores for rows which correlate to more other rows.

Since the ECDF curve monotonically increases from 0 to 1, for intervals with fixed
width, they contribute more to the ATC scores if they are close to 0 (Figure 3).

```{r, echo = FALSE, fig.width = 5, fig.height = 5, fig.align = "center", fig.cap = "Figure 3. Fixed-width intervals on the ECDF"}
par(mar = c(4, 4, 1, 1))
cor_v = c(0.00296, 0.00323, 0.00493, 0.00511, 0.00513, 0.00653, 0.00838, 0.00859, 0.00923, 0.0116, 0.0141, 0.0142, 0.0144, 0.0161, 0.0186, 0.0195, 0.0205, 0.0222, 0.0231, 0.0237, 0.0242, 0.025, 0.0255, 0.0277, 0.028, 0.029, 0.029, 0.0314, 0.0322, 0.0355, 0.0357, 0.0363, 0.0379, 0.0392, 0.0411, 0.0423, 0.0426, 0.0441, 0.0449, 0.046, 0.0501, 0.0508, 0.0529, 0.0536, 0.054, 0.0541, 0.0549, 0.0567, 0.0576, 0.0581, 0.0605, 0.0608, 0.0639, 0.0671, 0.0682, 0.0698, 0.0717, 0.0737, 0.0747, 0.075, 0.0773, 0.0816, 0.0834, 0.0837, 0.0869, 0.0899, 0.0928, 0.0937, 0.0945, 0.0946, 0.0964, 0.0981, 0.099, 0.101, 0.102, 0.102, 0.103, 0.104, 0.106, 0.109, 0.111, 0.115, 0.115, 0.116, 0.116, 0.121, 0.124, 0.127, 0.127, 0.127, 0.133, 0.135, 0.135, 0.135, 0.138, 0.143, 0.144, 0.146, 0.156, 0.16, 0.164, 0.169, 0.173, 0.178, 0.179, 0.179, 0.194, 0.202, 0.211, 0.253, 0.396, 0.429, 0.447, 0.455, 0.457, 0.465, 0.466, 0.477, 0.487, 0.488, 0.498, 0.518, 0.519, 0.522, 0.524, 0.526, 0.528, 0.539, 0.54, 0.541, 0.544, 0.546, 0.548, 0.551, 0.552, 0.556, 0.558, 0.559, 0.565, 0.568, 0.568, 0.574, 0.574, 0.576, 0.576, 0.579, 0.58, 0.582, 0.585, 0.59, 0.591, 0.592, 0.594, 0.596, 0.597, 0.609, 0.614, 0.615, 0.628)

cor_v = sort(cor_v)
f = ecdf(cor_v)
x = seq(0, 1, length = 100)
y = f(x)

plot(NULL, xlim = c(0, 1), ylim = c(0, 1), xlab = "x = |correlation|", ylab = "P(X <= x)")
x2 = seq(0, 1, length = 100)
polygon(c(x2, rev(x2)), c(f(x2), rep(1, length(x2))), col = "#FF000040", border = NA)
polygon(c(x2, rev(x2)), c(f(x2), rep(0, length(x2))), col = "#00FF0040", border = NA)
lines(x, y)
abline(v = seq(0.05, 0.95, by = 0.05), lty = 3, col = "grey")
```

There can be scenarios when a huge number of rows correlate to each other but
only with very small correlation values. It results in small right-shift of
the ECDF curves for these rows, compared to the scenario where rows are
completely uncorrelated. Since the correlation values are close to zero, these
small shifts of the ECDF curves results in a relative large increase of ATC
scores.

To decrease such effect, the ATC definition can be modified to:

$$ATC_i = (1 - \alpha) - \int_{\alpha}^1F(y)dy$$

where $y = x^\beta$. Now $ATC_i$ is the red area above ECDF curve only on the right of $y =
\alpha$. The coefficient $\beta$ is the power added to the absolute
correlations that it decreases more for the smaller correlations. By Default
$\alpha$ is set to 0 and $\beta$ is set to 1.

We slightly change the previous simulation test that in the first group, we
set 500 rows with pairwise correlation to 0.1, which generates 560 rows.
Similarly, there are four plots illustrated in Figure 4.

```{r, echo = FALSE, fig.width = 14, fig.height = 14/4, fig.align = "center", fig.cap = "Figure 4. A similation test with 560 rows"}
set.seed(12345)
test_ATC(500, 10, 50, 0.1, 0.8, 0.5)
```

In Figure 4, we can see group 1 (the black dots), since there are quite a lot of rows, they
gain high ATC scores even when they only have tiny correlation values (compare 
to the rows in group 2, the red dots).

To remove the effect of small correlation, we can set a value larger than 0 to
$\alpha$, e.g. 0.3:

```{r, echo = FALSE, fig.width = 14, fig.height = 14/4, fig.align = "center", fig.cap = "Figure 5. A similation test, alpha = 0.3"}
set.seed(12345)
test_ATC(500, 10, 50, 0.1, 0.8, 0.5, min_cor = 0.3)
```

In Figure 5, now we see the ATC scores for rows in group 1 decrease.

We can also set $\beta$ to decrease the value for rows in group 1, as shown in Figure 6.

```{r, echo = FALSE, fig.width = 14, fig.height = 14/4, fig.align = "center", fig.cap = "Figure 6. A similation test, beta = 3"}
set.seed(12345)
test_ATC(500, 10, 50, 0.1, 0.8, 0.5, beta = 3)
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
